---
title: "The Hiatus Control Chart"
author: "Neil Pettinger"
date: "2025-01-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Step 0 - install/load the packages

I think we'll need the usual suspects...

```{r}
library(openxlsx)
library(scales)
library(tidyverse)
```

### Step 1 import the data

We'll use the Sep 2014 to Sep 2015 A&E dataset

```{r}
dataframe_01 <-
  read.xlsx("https://www.kurtosis.co.uk/data/01_ed_data.xlsx") |>  
  mutate(new_arrival_date_time = convertToDateTime(arrival_date_time)) |> 
  mutate(new_departure_date_time = convertToDateTime(departure_date_time)) |> 
  select(patient_id,
         new_arrival_date_time,
         new_departure_date_time,
         adm_ward)
```

### Step 2 - create arrival_week and los_mins variables 

```{r}
dataframe_01a <-
  dataframe_01 |> 
  mutate(arrival_week = floor_date(new_arrival_date_time,
                                   "weeks",
                                   1),
         los_mins = difftime(new_departure_date_time,
                             new_arrival_date_time,
                             units = "mins"))
```

### Step 3 - calculate the control chart variables for the first half of the year

```{r}
dataframe_01b_first <-
  dataframe_01a |> 
  filter(arrival_week <= '2015-03-23') |> 
  group_by(arrival_week) |> 
  summarize(no_of_attendances = n())

#Calculate the mean and limits
  first_mean <- mean(dataframe_01b_first$no_of_attendances) 
  first_mR <- mean(abs(diff(dataframe_01b_first$no_of_attendances))) 
  first_lcl <- first_mean - 2.66 * the_mR 
  first_ucl <- first_mean + 2.66 * the_mR
```

### Step 4 - calculate the control chart variables for the second half of the year

```{r}
dataframe_01b_second <-
  dataframe_01a |> 
  filter(arrival_week > '2015-03-23') |> 
  group_by(arrival_week) |> 
  summarize(no_of_attendances = n())

#Calculate the mean and limits
  second_mean <- mean(dataframe_01b_second$no_of_attendances) 
  second_mR <- mean(abs(diff(dataframe_01b_second$no_of_attendances))) 
  second_lcl <- second_mean - 2.66 * the_mR 
  second_ucl <- second_mean + 2.66 * the_mR
```

### Step 5 - attempt the whole thing

```{r}
dataframe_01c <-
  dataframe_01b |> 
  mutate(first_activity_line = if_else(arrival_week <= '2015-03-23',
                             no_of_attendances,
                             NA)) |> 
  mutate(second_activity_line = if_else(arrival_week > '2015-03-23',
                             no_of_attendances,
                             NA)) |> 
  mutate(first_mean_line = if_else(arrival_week <= '2015-03-23',
                             first_mean,
                             NA)) |> 
  mutate(second_mean_line = if_else(arrival_week > '2015-03-23',
                             second_mean,
                             NA)) |> 
  mutate(first_lcl_line = if_else(arrival_week <= '2015-03-23',
                             first_lcl,
                             NA)) |> 
  mutate(second_lcl_line = if_else(arrival_week > '2015-03-23',
                             second_lcl,
                             NA)) |> 
  mutate(first_ucl_line = if_else(arrival_week <= '2015-03-23',
                             first_ucl,
                             NA)) |> 
  mutate(second_ucl_line = if_else(arrival_week > '2015-03-23',
                             second_ucl,
                             NA))
```

### Step 6 - plot the graph

```{r warning = FALSE}
ggplot(data = dataframe_01c) +
  geom_point(aes(x = arrival_week,
                 y = first_activity_line)) +
  geom_point(aes(x = arrival_week,
                 y = second_activity_line)) +
  geom_line(aes(x = arrival_week,
                y = first_activity_line)) +
  geom_line(aes(x = arrival_week,
                y = second_activity_line)) +
  geom_line(aes(x = arrival_week,
                y = first_mean_line)) +
  geom_line(aes(x = arrival_week,
                y = second_mean_line)) +
  geom_line(aes(x = arrival_week,
                y = first_lcl_line),
            linetype = "dashed") +
  geom_line(aes(x = arrival_week,
                y = second_lcl_line),
            linetype = "dashed") +
  geom_line(aes(x = arrival_week,
                y = first_ucl_line),
            linetype = "dashed") +
  geom_line(aes(x = arrival_week,
                y = second_ucl_line),
            linetype = "dashed") +
  geom_ribbon(aes(arrival_week, 
                  ymax=`second_ucl_line`, ymin=`second_lcl_line`), 
              fill="green", 
              alpha=0.25) +
  geom_vline(xintercept = as.POSIXct('2015-03-26 12:00'),
             colour = "grey") +
  scale_y_continuous(limits = c(0, 1400),
                     breaks = seq(0, 1400, 200),
                     labels = comma) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line.x = element_line(color="grey"),
        axis.line.y = element_line(color="grey"),) +
  labs(title = "The Green Zone",
         subtitle = "Attendances per week in a year of two halves",
         x = element_blank(),
         y = element_blank())
  
```

