---
title: "Frequency Table using cut()"
author: "Neil Pettinger"
date: "2025-01-15"
output:
  html_document:
  theme: null
---

### Step 0 - install/load the packages

The file we're going to use is an .xlsx file so we'll need {openxlsx} and {scales} as well as the {tidyverse} suite of packages - though I think we'll only need {dplyr} and {ggplot2} from that suite.

```{r message = FALSE, warning = FALSE}
library(openxlsx)
library(tidyverse)
library(scales)
```

### Step 1 - import the data

The raw data file is available for download on the Kurtosis website. I find that with {openxlsx}'s read.xlsx() function, I need to fix the datetimes using the convertToDateTime() function from the same package.

```{r}
dataframe_01 <-
  read.xlsx("https://www.kurtosis.co.uk/data/01_ed_data.xlsx") |>  
  mutate(new_arrival_date_time = convertToDateTime(arrival_date_time)) |> 
  mutate(new_departure_date_time = convertToDateTime(departure_date_time)) |> 
  select(patient_id,
         new_arrival_date_time,
         new_departure_date_time,
         adm_ward)
```

### Step 2 - create a los_mins variable

I'm using difftime() to do the subtraction arithmetic to get my length of stay variable, but I'm also applying as.numeric() to it to make sure it behaves like a proper number!

I'm also filtering out the three attendances that had length of stay of either less than zero or more than 1440.

```{r}
dataframe_01a <-
  dataframe_01 |>
  mutate(los_mins = as.numeric(difftime(new_departure_date_time,
                             new_arrival_date_time,
                             units = "mins"))) |> 
  filter(los_mins >=0) |> 
  filter(los_mins <= 1440)
```


### Step 2 - use cut() to create bins

I think the trick with cut() is that you have to have one fewer labels than breaks. If, in this code chunk, I'd written labels = seq(0, 1440, 60), I'd have got an error (number of intervals and length of 'labels' differ):

```{r}
dataframe_01b <-
  dataframe_01a |>
  mutate(los_cat = cut(los_mins,
                       breaks = seq(0, 1440, 60),
                       labels = seq(60, 1440, 60),
                       include.lowest = TRUE)) |> 
  arrange(desc(los_mins))
```

### Step 3 - can we summarize this?

This is a fairly straightforward group_by() and summarize() bit.

```{r}
dataframe_01c <-
  dataframe_01b |> 
  group_by(los_cat) |> 
  summarize(no_of_attendances = n())
```

### Step 4 - can we plot this?

One thing I was a bit worried about was whether my cut()-generated variable - los_cat - would cause problems for {ggplot2}. That variable is a factor variable. But I think as long as I keep it as my x-axis variable, I think I'm OK.

```{r}
ggplot(data = dataframe_01c) +
  aes(x = los_cat,
      y = no_of_attendances) +
  geom_col()
```

