---
title: "group_by() will never work"
author: "Neil Pettinger"
date: "2025-01-16"
output: html_document
---

### The frequency table problem (continued...)
The real problem I want to solve here is the problem of bins that contain zero values. I want my summary table to show that there are zero values in that bin, I don't want the bin to just not be there. That's no good. The zero-value-ness of the empty bin is an important piece of data.

And I;ve finally reached the realisation that you cannot therefore use group_by() in a way to get to such a frequency table. The only way you can do it is by specifyn gthe bins ins an empty bin dataframe and then ppulating the bins in turn.

Let's give it a go, I'll have the raw data - which I'll modifiy a bit so that it definitley has empty categories - maybe just do it on the basis of a week;s data.

### Step 0 - install/load the packages

```{r message = FALSE, warning = FALSE}
library(openxlsx)
library(tidyverse)
library(scales)
```


### Step 1 - import the data

```{r}
dataframe_01 <-
  read.xlsx("https://www.kurtosis.co.uk/data/01_ed_data.xlsx") |>  
  mutate(new_arrival_date_time = convertToDateTime(arrival_date_time)) |> 
  mutate(new_departure_date_time = convertToDateTime(departure_date_time)) |> 
  select(patient_id,
         new_arrival_date_time,
         new_departure_date_time,
         adm_ward) |> 
  filter(new_arrival_date_time >= '2014-09-29 00:00') |> 
  filter(new_arrival_date_time <= '2014-10-05 23:59')
```

### Step 2 - create the los_mins variable

```{r}
dataframe_01a <-
  dataframe_01 |>
  mutate(los_mins = as.numeric(difftime(new_departure_date_time,
                             new_arrival_date_time,
                             units = "mins"))) |> 
  filter(los_mins >=0) |> 
  filter(los_mins <= 1440)
```


### Step 3 - check to see if there are 'missing' values (or empty bins)

```{r}
dataframe_01b <-
  dataframe_01a |>
  mutate(los_cat = cut(los_mins,
                       breaks = seq(0, 1440, 60),
                       labels = seq(60, 1440, 60),
                       include.lowest = TRUE)) |> 
  arrange(desc(los_mins))
```

Next:

```{r}
dataframe_01c <-
  dataframe_01b |> 
  group_by(los_cat) |> 
  summarize(no_of_attendances = n())
```
 Yes, there are two empty bins - the 1141-1200 and the 1381-1440 bins
 
 So: to address this, I need to first of all create a dataframe that contains these upper and lower bin values
 
```{r}
dataframe_bins <-
  data.frame(bin_lower = seq(0, 1380, 60),
             bin_upper = seq(60, 1440, 60))
```
 
And now I need to populatue this datafram directly using a left_join to dataframe_01a
```{r}
dataframe_01d <-
  dataframe_01a |> 
  left_join(dataframe_bins,
            join_by(los_mins > bin_lower,
                    los_mins <= bin_upper)) |> 
  arrange(los_mins)
  
```
 OK, do a group_by()
 
```{r}
dataframe_01e <-
  dataframe_01d |> 
  group_by(bin_upper) |> 
  summarize(no_of_attendances = n())
```
 And now join this summary table to the dataframe_bins dataframe to get the emtpy bins
 
```{r}
dataframe_01f <-
  dataframe_bins |> 
  left_join(dataframe_01e,
            join_by(bin_upper == bin_upper)) |> 
  mutate(new_no_of_attendances = if_else(is.na(no_of_attendances),
                                         0,
                                         no_of_attendances)) |> 
  select(bin_upper,
         new_no_of_attendances)
  
```
 
Can I draw a bar chart with this daat - even though there are NA values?

```{r}
ggplot(data = dataframe_01f) +
  aes(x = bin_upper,
      y = new_no_of_attendances) +
  geom_col() +
  scale_x_continuous(breaks = seq(60, 1440, 60)) +
  theme_classic()
```

