---
title: "Frequency Table using left_join()"
author: "Neil Pettinger"
date: "2025-01-14"
output: html_document
---

### Step 0 - install/load the packages

The file we're going to use is an .xlsx files so we'll need {openxlsx} as well as the {tidyverse} suite of packages - though I think we'll only need {dplyr} and {ggplot2} from that suite.

```{r message = FALSE}
library(openxlsx)
library(tidyverse)
library(scales)
```

### Step 1 - import the data

The raw data file is available for download on the Kurtosis website. I find that with {openxlsx}'s read.xlsx() function, I need to fix the datetimes using the convertToDateTime() function from the same package.

```{r}
dataframe_01 <-
  read.xlsx("https://www.kurtosis.co.uk/data/01_ed_data.xlsx") |>  
  mutate(new_arrival_date_time = convertToDateTime(arrival_date_time)) |> 
  mutate(new_departure_date_time = convertToDateTime(departure_date_time)) |> 
  select(patient_id,
         new_arrival_date_time,
         new_departure_date_time,
         adm_ward)
```

### Step 2 - create a los_mins variable

I'm using difftime() to do the subtraction arithmetic to get my length of stay variable, but I'm also applying as.numeric() to it to make sure it behaves like a proper number!

I'm also filtering out the three attendances that had length of stay of either less than zero or more than 1440.

```{r}
dataframe_01a <-
  dataframe_01 |>
  mutate(los_mins = as.numeric(difftime(new_departure_date_time,
                             new_arrival_date_time,
                             units = "mins"))) |> 
  filter(los_mins >=0) |> 
  filter(los_mins <= 1440)
```

# Step 3 - create a dataframe of bins

Meanwhile, we create a dataframe that contains the upper and lower limits of the bins we want to summarize the raw data into.

```{r}
dataframe_bins <-
  data.frame(bin_lower = seq(0, 1380, 60),
             bin_upper = seq(60, 1440, 60))
```

#Step 4 - Do a left_join()

Having got the 24x2 bins dataframe in place, we can now attempt a left_join() as our first step to creating the frequency table.

```{r}
dataframe_01b <-
  dataframe_01a |> 
  left_join(dataframe_bins,
            join_by(los_mins >= bin_lower,
                    los_mins < bin_upper))
```

#Step 5 - Now need to group this dataframe...

If we now group this dataframe by the bin_upper variable, we should get a neat summary table.

```{r}
dataframe_01c <-
  dataframe_01b |> 
  group_by(bin_upper) |> 
  summarize(no_of_attendances = n())
```


#Step 6 - draw a histogram using geom_col()

```{r}
ggplot(data = dataframe_01c) +
  aes(x = bin_upper,
      y = no_of_attendances) +
  geom_col(width = 60,
           colour = "white",
           fill = "blue") +
  scale_x_continuous(limits = c(0, 1440),
                     breaks = seq(60, 1440, 60)) +
  scale_y_continuous(labels = comma) +
  labs(title = "The Frequency Distribution",
       subtitle = "Anytown General Hospital",
       caption = "Source: Patient Management System",
       x = "Length of stay (minutes)",
       y = element_blank()) +
  theme_minimal()
```

